\documentclass{beamer}

\newcommand{\course}{CS 1331 Introduction to Object Oriented Programming}
\newcommand{\lesson}{Arrays, Part 1 of 3}
\newcommand{\code}{http://www.cc.gatech.edu/~simpkins/teaching/gatech/cs1331/code}

\author[Chris Simpkins] 
{Christopher Simpkins \\\texttt{chris.simpkins@gatech.edu}}
\institute[Georgia Tech] % (optional, but mostly needed)

\date[CS 1331]{}
\subject{\lesson}
% This is only inserted into the PDF information catalog. Can be left
% out. 

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% \pgfdeclareimage[width=0.6in]{coc-logo}{cc_2012_logo}
% \logo{\pgfuseimage{coc-logo}}

\mode<presentation>
{
  \usetheme{Berlin}
  \useoutertheme{infolines}

  % or ...

 \setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}

\usepackage{hyperref}
\usepackage{fancybox}
\usepackage{listings}
\usepackage[abbr]{harvard}

\usepackage[english]{babel}
% or whatever

\usepackage[latin1]{inputenc}
% or whatever

\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.


\usepackage{listings}
 
% "define" Scala
\lstdefinelanguage{scala}{
  morekeywords={abstract,case,catch,class,def,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,val,var,while,with,yield},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]""",
}

\usepackage{color}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}
 
% Default settings for code listings
\lstset{frame=tb,
  language=scala,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\scriptsize\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  frame=single,
  breaklines=true,
  breakatwhitespace=true,
  keepspaces=true
  %tabsize=3
}


\title[\course] % (optional, use only with long
                                      % paper titles)
{\lesson}

\subtitle{}
%% {Include Only If Paper Has a Subtitle}


% Delete this, if you do not want the table of contents to pop up at
% the beginning of each subsection:
%% \AtBeginSection[]
%% {
%%   \begin{frame}<beamer>{Outline}

%%  \tableofcontents[currentsection]
%%   \end{frame}
%% }

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command: 

% \beamerdefaultoverlayspecification{<+->}


\begin{document}

\begin{frame}
  \titlepage
\end{frame}


%------------------------------------------------------------------------
\begin{frame}[fragile]{Arrays}


Java Arrays (\href{http://docs.oracle.com/javase/specs/jls/se7/html/jls-10.html}{JLS \S 10}):
\begin{itemize}
\item are objects,
\item are dynamically allocated (e.g., with operator {\tt new}), and
\item have a fixed number of elements of the same type.
\end{itemize}



\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Creating Arrays}


Consider the following {\it array creation expression} (\href{http://docs.oracle.com/javase/specs/jls/se7/html/jls-10.html#jls-10.3}{JLS \S 10.3)}:
\begin{lstlisting}[language=Java]
double[] scores = new double[5];
\end{lstlisting}
This declaration:
\begin{itemize}
\item allocates a 5-element array,
\item the {\tt 5} in the example above can be any expression that is unary promotable to an {\tt int} (\href{http://docs.oracle.com/javase/specs/jls/se7/html/jls-5.html#jls-5.6.1}{JLS \S 5.6.1})
\item stores the address of this new array in {\tt scores}, and
\item initializes each value to it's default value ({\tt 0} for numeric types, {\tt false} for {\tt boolean} types, and {\tt null} for references, \href{http://docs.oracle.com/javase/specs/jls/se7/html/jls-4.html#jls-4.12.5}{JLS \S 4.12.5}).
\end{itemize}


\end{frame}
%------------------------------------------------------------------------



%------------------------------------------------------------------------
\begin{frame}[fragile]{Array Declarations}


The preceding array definition
\begin{lstlisting}[language=Java]
double[] scores = new double[5];
\end{lstlisting}
could be split into a declaration and initialization:
\begin{lstlisting}[language=Java]
double[] scores;
scores = new double[5];
\end{lstlisting}
Also, you can put the {\tt []} on the type or the variable name when delaring an array.  These two declarations are equivalent:
\begin{lstlisting}[language=Java]
double[] scores; 
double scores[];
\end{lstlisting}
Generally, it's better style to put the {\tt []} on the type.


\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Mixed Declarations}


Note that you can mix aray declarations with declarations of variables having the same element type.  The declaration line:
\begin{lstlisting}[language=Java]
double scores[], average;
\end{lstlisting}
creates
\begin{itemize}
\item an uninitialzed array of {\tt double} named {\tt scores}, and
\item a {\tt double} variable named {\tt average} (auto-initialized to the default value for {\tt double}, namely, {\tt 0.0} if it's an instance variable).
\end{itemize}

What is the value of {\tt scores} after the statement above?

\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Array Objects}


After the definition:
\begin{lstlisting}[language=Java]
double[] scores = new double[5];
\end{lstlisting}
{\tt scores} points to an array object in memory that can be visualized as:\\
\vspace{.1in}
\begin{tabular}{p{.5in}p{.5in}p{.5in}p{.5in}p{.5in}}
0 & 1 & 2 & 3 & 4 \\
\end{tabular}
\begin{tabular}{|p{.5in}|p{.5in}|p{.5in}|p{.5in}|p{.5in}|}\hline
0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\
\hline
\end{tabular}\\
\vspace{.1in}
The {\it indexes} of {\tt scores} range from 0 to 4.  The size of arrays are stored in a {\tt public final} instance variable named {\tt length}

\begin{lstlisting}[language=Java]
scores.length == 5;
\end{lstlisting}
What is the type and value of the expression above?

\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Accessing Array Elements}


Array elements are accessed with an {\tt int}-promotable expression enclosed in square brackets ({\tt []})
\begin{lstlisting}[language=Java]
double[] scores = new double[5];
scores[0] = 89;
scores[1] = 100;
scores[2] = 95.6;
scores[3] = 84.5;
scores[4] = 91;
scores[scores.length - 1] = 99.2;
\end{lstlisting}


\vspace{.1in}
Will this line compile?  If so, what will happen at runtime?
\begin{lstlisting}[language=Java]
scores[scores.length] = 100;
\end{lstlisting}


\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Initializing Arrays}


You can provide initial values for (small) arrays 
\begin{lstlisting}[language=Java]
String[] validSuits = {"diamonds", "clubs", "hearts","spades"};
\end{lstlisting}

\begin{itemize}
\item What is {\tt validSuits.length}?
\item What is {\tt validSuits[1]}?
\end{itemize}

You can also use a loop to initialize the values of an array:
\begin{lstlisting}[language=Java]
int[] squares = new int[5];
for (int i = 0; i < squares.length; ++i) {
    squares[i] = i*i;
}
\end{lstlisting}

What is {\tt squares[4]}? 

\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Traversing Arrays}


Arrays and {\tt for} statements go hand-in-hand:
\begin{lstlisting}[language=Java]
double[] scores = new double[5];
for (int i = 0; i < 5; ++i) {
    System.out.printf("scores[%d] = %.2f%n", i, scores[i]);
}    
\end{lstlisting}

You can also use the ``enhanced'' for loop:

\begin{lstlisting}[language=Java]
for (double score: scores) {
    System.out.println(score);
}
\end{lstlisting}

Read the enhanced for loop as ``for each element of the array ...''.

\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Traditional {\tt for} Versus {\tt for}-each}


In cases where you don't need the index, use the enhanced for loop.  Consider:
\vspace{-.05in}
\begin{lstlisting}[language=Java]
double sum = 0.0;
for (int i = 0; i < scores.length; ++i) {
    sum += scores[i];
} 
\end{lstlisting}
\vspace{-.05in}
In the code above, {\tt scores.length} is used only for bounding the array traversal, and the index {\tt i} is only used for sequential array access.  Those are two things we can mess up.  The enhanced for loop is cleaner:

\begin{lstlisting}[language=Java]
double sum = 0.0;
for (double score: scores) {
    sum += score;
} 
\end{lstlisting}

Also note how our naming conventions help to make the code clear.  You can read the loop above as ``for each score in scores''.

\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Array Initialization and Access Gotchas}


Because arrays are allocated dynamically, this will compile:
\begin{lstlisting}[language=Java]
double[] scores = new double[-5];
\end{lstlisting}
but will produce an error at run-time:
\begin{lstlisting}[language=Java]
Exception in thread "main" java.lang.NegativeArraySizeException
        at ArrayBasics.main(ArrayBasics.java:4)
\end{lstlisting}

Also, array access expressions are evaluated and checked at run-time.  So, in the same way that accessing an array with an index $\ge$ the size of the array produces a run-time error, negative indexes like:

\begin{lstlisting}[language=Java]
scores[-1] = 100;
\end{lstlisting}
produce:
\begin{lstlisting}[language=Java]
Exception in thread "main" java.lang.ArrayIndexOutOfBoundsException: -1
        at ArrayBasics.main(ArrayBasics.java:23)
\end{lstlisting}
\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Arrays as Method Parameters - {\tt main}}


We've already seen an array parameter:
\begin{lstlisting}[language=Java]
public static void main(String[] args)
\end{lstlisting}

We can use this array just like we use any other array.
\begin{lstlisting}[language=Java]
public class Shout {

    public static void main(String[] args) {
        for (String arg: args) {
            System.out.print(arg.toUpperCase() + " ");
        }
        System.out.println();
    }
}
\end{lstlisting}
See also \href{\code/CourseAverage.java}{CourseAverage.java}
\end{frame}
%------------------------------------------------------------------------

%------------------------------------------------------------------------
\begin{frame}[fragile]{Passing and Returning Arrays}


Arrays are objects, so they can be passed to methods and returned from methods just like other objects.\\
\vspace{.1in}
Consider the following Array copy method:
\begin{lstlisting}[language=Java]
public static int[] copyOf(int[] sourceArray) {
    int[] newArray = new int[sourceArray.length];
    for (int i = 0; i < newArray.length; ++i) {
      newArray[i] = sourceArray[i];
    }
    return newArray;
}
\end{lstlisting}
The array returned from {\tt copyOf} is value-equivalent to the array passed to the method, but they are different arrays.  Modifying the returned array does not modify the passed-in array.


\end{frame}
%------------------------------------------------------------------------


%% %------------------------------------------------------------------------
%% \begin{frame}[fragile]{Arrays}


%% \begin{lstlisting}[language=Java]

%% \end{lstlisting}

%% \begin{itemize}
%% \item
%% \end{itemize}


%% \end{frame}
%% %------------------------------------------------------------------------

% %------------------------------------------------------------------------
% \begin{frame}[fragile]{}


% \begin{lstlisting}[language=Java]

% \end{lstlisting}

% \begin{itemize}
% \item
% \end{itemize}


% \end{frame}
% %------------------------------------------------------------------------


\end{document}
